### 2020-09-08 paper/paper.tex
edit intro

Across the world, vulnerable populations rely on governments for access to basic public services such as education. Over the past two decades, enrollment rates in public education have improved dramatically.\footnote{See \citet{haggard_development_2008}, \citet{kaufman_globalization_2001}.} Improvements in access have not been matched, for many, with a better quality of education. From a global perspective, educational quality indicators such as the PISA (Programme for International Student Assessment) score have broadly stagnated.\footnote{For coverage, see [here](https://www.economist.com/international/2019/12/05/pisa-results-can-lead-policymakers-astray).}

How can we improve the quality of education? In large part, the answers are domestic. While each country faces a particular set of challenges, broad institutional reforms in the developing world make their cases comparable. Decentralization has delegated responsibility for public education to subnational governments, as well as induced variation in the quality of education along subnational lines \citep{falleti_decentralization_2010, arretche_mitos_1996}. Understanding how this level of government manages these services, and in particular how local actors strategically shape them can help find answers to that question. \citep{gulzar_politicians_2017, min_power_2015}.

% In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services. Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.^[@laver_coalitions_1990, @power_optimism_2010.] As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service (@ingersoll_teacher_2001, @ronfeldt_how_2013). Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.

% To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education, both at a national and subnational level. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory, and a school-grade-specific turnover index. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

@@

% This study contributes to an emerging literature on the politics of personnel and public services (@pepinsky_bureaucracy_2017, @finan_personnel_2015 and @gulzar_politicians_2017). Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (@ferraz_corrupting_2012). It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (@kohli_state-directed_2004, @evans_embedded_1995).

@@

% Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (@centeno_unpacking_2017, @kohli_state-directed_2004). A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (@johnson_miti_1982, kohli_state-directed_2004). A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (@evans_bureaucracy_1999).

% Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. @gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. @akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

@@

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (@grindle_jobs_2012, @mccarty_appointments_2004). Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. (@raile_executive_2011, @power_optimism_2010).

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. (@raile_executive_2011). In a setting characterized by weak party cohesion and programmatic commitments (@ames_electoral_1995, @lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (@batista_o_2013, @neto_presidential_2006, @figueiredo_executivo_1999).

@@

% Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (@gailmard_slackers_2007). Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (@clotfelter_teacher_2007.) @akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (@colonnelli_patronage_2017).

@@

% I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power. In particular, mayors are assumed to be career-oriented and re-election seeking (@besley_principled_2006). This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (@stokes_brokers_2013, @robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.

@@

% The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors. The intuition is as follows. Mayors need to secure a simple majority of legislative votes to have the budget approved. @groseclose_buying_1996 and @banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor. Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage. The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.

% The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures. Starting with @barro_control_1973 and @ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (@coate_form_1995; @besley_principled_2006). The possibility of reelection can motivate them to exert effort and convince voters to reelect them. This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (@groseclose_buying_1996, @banks_buying_2000).

@@

% I introduce a set of simplifying assumptions. First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so. Note that this is an informational heuristic. Voters have beliefs about how competent a politician is and strictly prefer them over the alternative. Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (@coate_form_1995). Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations. Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.

% The game has two periods. Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$. An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy. The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See @kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.

@@

% The mayor's choice set is the bureaucratic quality parameter $\theta_t$. As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature. This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year. Adapting @gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic. Technocratic politicians invariably deliver high quality public goods to the voter. Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.

@@

% In Brazil, responsibility for primary education is delegated to local governments (@paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

@@

% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (@brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (@gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (@militao_educacao_2014).

% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (@ferraz_corrupting_2012).

@@

% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by @souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (@dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

@@

% I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by @pereira_junior_indicadores_2016. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

@@

% For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$. This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (@diprete_multilevel_1994, @lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise. The main specification is as follows:
% \begin{align*}
% \text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
% \psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
% \end{align*}
% We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover. In this set-up, $\beta_3$ is the additional effect of staff turnover on test scores if students are in 9th grade. The first level is the grade, with associated characteristics $X_isjt$ at the grade level: among others, percentage of students who have not graduated last year and share of children with a fridge in their house. The second is the school $s$, with characteristics $V_{sjt}$, such as access to internet or the presence of a cafeteria for students. The third level is the municipality $j$, including municipal sociodemographic characteristics ($W_{jt}$) such as population size and median wages, as well as per pupil budgetary expenditures on eduation. Finally, I include state ($\alpha_kj$) and year ($\delta_t$) fixed effects to account for unobserved time-invariant state characteristics and annual seasonality.

@@

% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in @imai_covariate_2014 which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

% This bureaucratic turnover has important consequences for the quality of education delivered in municipalities. I find that turnover has a negative effect on student learning, across a set of specifications for turnover and different evaluation metrics for student learning. The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning. This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (@gottlieb_countervailing_2019).

>>>>

% Across the world, vulnerable populations rely on governments for access to basic public services such as education. In many countries, particularly in the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, mathematical operations, elusive. Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them. \citep{cochrane_policies_1983, falleti_decentralization_2010}. How can local governments improve the quality the of public services they deliver? In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?

In this paper, I highlight the role of the strategic interaction between political elites and its downstream effects on the provision of educational services. Similar to coalition building, local executive leaders secure broader support for their policy agenda through the allocation of public sector positions to city councilors.\citep{laver_coalitions_1990, power_optimism_2010} As highlighted by previous research, teacher turnover can lead to decreased morale and inexperienced staff entering public service \citep{ingersoll_teacher_2001, ronfeldt_how_2013}. Patronage, by inducing new hires or dismissals, can therefore cause instability in the educational sector, with negative effects for student learning.

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education, both at a national and subnational level. I construct a multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory, and a school-grade-specific turnover index. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

@@

% This study contributes to an emerging literature on the politics of personnel and public services (pepinsky_bureaucracy_2017, finan_personnel_2015 and gulzar_politicians_2017). Focusing on political actors refines our understanding of local politics and how intra-elite bargain reshape bureaucracies, leading to welfare loss for the broader population, as subnational actors divert resources from public services (ferraz_corrupting_2012). It also shifts focus from provision to the implementation of public services, echoing a long-standing literature on state capacity and the necessity for bureaucratic autonomy (kohli_state-directed_2004, evans_embedded_1995).

@@

% Bureaucracies have a clear impact on the delivery of public services. A long-standing literature on state capacity provides a theoretical and substantive foundation to analyze bureaucratic institutions (centeno_unpacking_2017, kohli_state-directed_2004). A first generation of scholars, focusing on the successful developmental cases of East Asia, highlighted the need for a technocratic and autonomous bureaucracy (johnson_miti_1982, kohli_state-directed_2004). A Weberian wall separating bureaucrats from elected officials was considered indispensable for the successful provision of economic growth (evans_bureaucracy_1999).

% Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. gulzar_politicians_2017 show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. akhtari_political_2015, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

@@

% For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors (grindle_jobs_2012, mccarty_appointments_2004). Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. (raile_executive_2011, power_optimism_2010).

% In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. (raile_executive_2011). In a setting characterized by weak party cohesion and programmatic commitments (ames_electoral_1995, lucas_ideological_2010), bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda (batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999).

@@

% Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population (gailmard_slackers_2007). Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher (clotfelter_teacher_2007.) akhtari_political_2015 finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones (colonnelli_patronage_2017).

@@

% I model the executive decision over patronage as a product of political career concerns and strategic interaction with the local legislative power. In particular, mayors are assumed to be career-oriented and re-election seeking (besley_principled_2006). This provides an alternative set-up to extant treatments of patronage as a pre-electoral bargain, naturally highlighting the question of government performance (stokes_brokers_2013, robinson_political_2013.) The key feature of the model is the trade-off between patronage and quality of public services, which mayors are aware of when allocating public sector jobs to city councilors.

@@

% The main proposition is that mayors who control a smaller (larger) share of legislative seats allocate more (less) public sector job to city councilors. The intuition is as follows. Mayors need to secure a simple majority of legislative votes to have the budget approved. groseclose_buying_1996 and banks_buying_2000 point out that the executive should prioritize votes from cheaper legislators, which in this set-up are those who share the same electoral coalition as the mayor. Having exhausted these votes, the executive has to resort to buying votes from the opposition, which demand more patronage. The less seats occupied by members of the mayor's electoral coalition, the more expensive vote-buying becomes, and the observable implication is increased patronage in the municipality.

% The model builds on rich theoretical treatment of electoral accountability and majority-buying in legislatures. Starting with barro_control_1973 and ferejohn_incumbent_1986, political economists have proposed formal models analyzing the nexus of electoral accountability as an incentive for shaping the behavior of elected politicians (coate_form_1995; besley_principled_2006). The possibility of reelection can motivate them to exert effort and convince voters to reelect them. This model extends previous analyses by incorporating legislators to the game, with the need to secure a legislative majority to implement the executive agenda (groseclose_buying_1996, banks_buying_2000).

@@

% I introduce a set of simplifying assumptions. First, mayors are one of two types: a technocratic politician who delivers high quality public goods, and a pragmatic one, who fails to do so. Note that this is an informational heuristic. Voters have beliefs about how competent a politician is and strictly prefer them over the alternative. Because voters cannot observe the type, they update beliefs observing the quality of public goods they observe (coate_form_1995). Additionally, I assume that the voter is sincere, excluding pivotality concerns and strategic voting from voters' utility calculations. Finally, I normalize the politician's exit option to 0: the politician always prefers to be re-elected.

% The game has two periods. Voter $v$ does not observe the amount of patronage handed out by mayor $m$ nor her type, but observes and derives utility from the realization of public good quality $\omega_t$. An important feature in the model is that high quality public services $\omega_{h}$ are produced probabilistically as a function of $\theta$, the quality of the bureaucracy. The more jobs a mayor allocates to legislators, the lower is $\theta$, thus capturing the inverse relation between patronage and state capacity.^[See kohli_state-directed_2004.] The budgetary process also affects the public service delivery, with a failure to approve the budget lowering the resources available to the executive.

@@

% The mayor's choice set is the bureaucratic quality parameter $\theta_t$. As quality increases - the mayor allocates a smaller proportion of jobs to legislators - the less votes become secure in the legislature. This reduces the likelihood of securing the necessary resources to provide public goods for a given fiscal year. Adapting gailmard_slackers_2007, there are two types of mayors: technocratic and pragmatic. Technocratic politicians invariably deliver high quality public goods to the voter. Pragmatic politicians, on the other hand, are susceptible to legislative pressure and at times fail to deliver them.

@@

% In Brazil, responsibility for primary education is delegated to local governments (paschoal_historia_2009). The municipal educational system has increased in relevance over the past few decades, as municipalities took on a larger proportion of students from state-level governments. Figure \ref{fig:enrollment} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

@@

% The local executive branch has sole prerogative over the hiring and firing of teachers and school principals. School principals are considered positions of trust (\textit{cargos de confiança}) and usually appointed by the mayor or secretary of education (brollo_political_2017). Teachers usually enter the public service through a public exam and are eligible for tenure after two years (gatti_formacao_2010) However municipalities increasingly resort to temporary contracts to hire new teachers as budgetary pressures amount. Municipal teachers and school principals are overseen by a local department of education (militao_educacao_2014).

% Funding for municipal education relies primarily on federal transfers, the Fundeb, derived from 10 percent of tax revenues for each level of government to education. At the municipal level, 25 percent of the local budget must be allocated to education, and compensatory federal transfers are institutionalized by law to those municipalities which do not reach the target.^[For more details on the Fundeb, see (here)[https://www.fnde.gov.br/index.php/financiamento/fundeb/sobre-o-plano-ou-programa/sobre-o-fundeb].] While municipalities may be audited to verify whether funds are being properly used, personnel decisions and daily operations are fully under municipal discretion (ferraz_corrupting_2012).

@@

% Brazil is comprised by over 5 thousand municipalities, each with their own city hall and council. Mayors and city councillors (\textit{vereadores}) are democratically elected every four years, with the possibility of reelection for both.^[The last municipal elections were in 2016. These take place ] The executive is responsible for the implementation of social policies such as education, as well as drafting the municipal budget. The city council, on the other hand, oversees legislation and approves the budget for the fiscal year. As noted by souza_reforma_2004, decisions over how to manage education are primarily in the hands of the mayor, but the city councilors exert significant pressure.

% In order to win elections and garner support for their campaign, mayors form electoral alliances with local party labels (dantas_eleicoes_2013). These mayoral coalitions, once in government, are an important basis for approving budgets and, more generally, executive control over the legislative chamber. Interviews with city councilors indicate that the legislature is divided into a pro-government (\textit{governo}) and opposition (\textit{oposição}) groups (\textit{bancada}). While these electoral coalitions do not necessarily remain intact once governments are formed, fragmentation is generally on the margins, with mayors swapping defecting legislators for new ones.^[Interview with C, former chief of staff of municipality J. September 2018.]

@@

% I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by pereira_junior_indicadores_2016. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

@@

% For the first estimation, I employ a hierarchical linear model to estimate the effect of teacher turnover on average test scores for grade $i$ at school $s$, in municipality $j$ and year $t$. This estimation strategy has been used in multiple studies to analyze educational, due to its natural multi-level setting (classroom, school and municipality) as well as ability to incorporate covariates at each level of the estimation (diprete_multilevel_1994, lee_using_2000.) Let $\text{grade}_{isjt}$ be a dummy variable equal to 1 if the classroom is in grade 9 and 0 otherwise. The main specification is as follows:
% \begin{align*}
% \text{test score}_{isjt} = \beta_1 \text{turnover}_{isjt} + \beta_2 \text{grade}_{isjt} +\beta_3 \text{turnover}_{isjt} \times \text{grade}_{isjt} + \\ 
% \psi X_{isjt} + \phi V_{sjt} +\zeta W_{jt}+ \alpha_k + \delta_t + \epsilon_{isjt}
% \end{align*}
% We are interested in $\beta_1$ and $\beta_3$, the change in test scores associated with teacher or school principal turnover. In this set-up, $\beta_3$ is the additional effect of staff turnover on test scores if students are in 9th grade. The first level is the grade, with associated characteristics $X_isjt$ at the grade level: among others, percentage of students who have not graduated last year and share of children with a fridge in their house. The second is the school $s$, with characteristics $V_{sjt}$, such as access to internet or the presence of a cafeteria for students. The third level is the municipality $j$, including municipal sociodemographic characteristics ($W_{jt}$) such as population size and median wages, as well as per pupil budgetary expenditures on eduation. Finally, I include state ($\alpha_kj$) and year ($\delta_t$) fixed effects to account for unobserved time-invariant state characteristics and annual seasonality.

@@

% Despite the use of different educational quality and staff turnover metrics, addition of a relatively extensive set of controls, the estimation strategy proposed here remains observational. In the absence of exogenous shock in either turnover or coalition shares, a credible identification strategy for each set of estimations remains a challenge. To partially address these concerns, as an additional robustness check I replicate these estimations with weights estimated through a covariate balancing propensity score outlined in imai_covariate_2014 which is generalized for continuous treatments. A covariate balance test for either estimation is presented in the appendix.

@@

% This bureaucratic turnover has important consequences for the quality of education delivered in municipalities. I find that turnover has a negative effect on student learning, across a set of specifications for turnover and different evaluation metrics for student learning. The evidence therefore seems to point out that mayors with a weaker hold on the city council resort to greater patronage, with negative consequences for student learning. This set of findings contribute to an emergent literature on the ambiguous consequences of stronger competition in weakly institutionalized contexts (gottlieb_countervailing_2019).
